---
title: "Data Types and Visualizations"
output: github_document
---

## Load packages

```{r}
library(tidyverse)
library(medicaldata) 
library(ggplot2)
```

## The data

```{r}
head(covid_testing)
```

```{r}
# To learn more about this dataset
?covid_testing
```

## CODING CHALLENGE 1

**Your turn!**

Identify the variable types for the following variables:
1. Subject ID, 
2. Age,
3. Gender,
4. Whether the specimen was collected via a drive-thru site, 
5. Cycle at which threshold reached during PCR 


If you're not sure what the column names are for those variables, check the [code book for this dataset](https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/codebooks/covid_testing_codebook.html))

```{r}
# Subject ID

# Age

# Gender

# Whether the specimen was collected via a drive-thru site

# Cycle at which threshold reached during PCR

```


** ---------------------------------------------- **
** STOP HERE!  Go back to the module to continue. **
** ---------------------------------------------- **

## Using `str()`

Now use `str()` to get the variable type for all of the variables in the dataset

```{r}
str(covid_testing)
```

## Example 1: Scatterplot (single color)

```{r}
baseplot <- ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result # y-axis variable
                             ))
baseplot
```


```{r}
baseplot + 
  geom_point() # adds the points to the scatterplot
```

### Changing the color of the points
```{r}
baseplot + 
  geom_point(  # adds the points to the scatterplot
    color = "red" # sets the color of the points to "red"
    )
```

## Example 2: Scatterplot (color by variable)

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = drive_thru_ind # color variable
                             )
       ) + 
  geom_point() # adds the points to the scatterplot
```

### Converting drive_thru_ind to a factor

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = as.factor(drive_thru_ind) # color variable
                             )
       ) + 
  geom_point() # adds the points to the scatterplot
```

### Changing the color of the points

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = as.factor(drive_thru_ind) # color variable
                             )
       ) + 
  geom_point() + # adds the points to the scatterplot
  scale_color_manual(values = c("0" = 'blue', # sets colors for each level
                                "1" = 'red'))
```

### Adding plot titles and labels

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = as.factor(drive_thru_ind)))+ # color variable 
  geom_point() + 
  scale_color_manual(values = c("0" = 'blue', # sets colors for each level
                                "1" = 'red')) + 
  ggtitle("CT Results by Pandemic Day") + # adds title to plot
  xlab("Pandemic Day") + # changes x-axis label
  ylab("CT Result") + # changes y-axis label
  labs(color = "Drive Thru Indicator") # changes legend label
```

### Changing plot placement and theme 

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = as.factor(drive_thru_ind))) + # color variable
  geom_point() + 
  scale_color_manual(values = c("0" = 'blue', # sets colors for each level
                                "1" = 'red')) + 
  ggtitle("CT Results by Pandemic Day") + # adds title to plot
  xlab("Pandemic Day") + # changes x-axis label
  ylab("CT Result") + # changes y-axis label 
  labs(color = "Drive Thru Indicator") +  # changes legend label
  theme_bw() + # changes plot theme to black and white
  theme(plot.title = element_text(hjust = 0.5), # centers plot title
        legend.position="bottom") # moves legend to bottom 
```

### Using facet_wrap

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             y = ct_result, # y-axis variable
                             color = as.factor(drive_thru_ind) # color variable
                             )
       ) + 
  geom_point() + 
  scale_color_manual(values = c("0" = 'blue', # sets colors for each level
                                "1" = 'red')) + 
  ggtitle("CT Results by Pandemic Day") + # adds title to plot
  xlab("Pandemic Day") + # changes x-axis label
  ylab("CT Result") + # changes y-axis label 
  labs(color = "Drive Thru Indicator") + # changes legend label
  theme_bw() + # changes plot theme to black and white
  theme(plot.title = element_text(hjust = 0.5), # centers plot title
        legend.position="bottom") +  # moves legend to bottom 
  facet_wrap(~gender)
```

## CODING CHALLENGE 2

**Your turn!** 

Create a scatterplot with "Time elapsed between receive time and verification time" on the y-axis and "Pandemic Day" on the x-axis. 

Additionally,
1. Color the points on the graph by demographic group. 
2. Add appropriate labels to the title, x-axis, y-axis, and legend.
3. Change the legend colors to "red","orange","green","blue", and "purple."
4. Put the legend on the left-hand side of the graph

Reminder: Codebook for the covid_testing dataset is available [here](https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/codebooks/covid_testing_codebook.html)

```{r}



```

** ---------------------------------------------- **
** STOP HERE!  Go back to the module to continue. **
** ---------------------------------------------- **

## Example 3: Histogram (color by variable)

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             fill = gender) # note: using "fill" not "color"
       ) + 
  geom_histogram(# adds the histograms to the graph
    bins = 10, # number of bins 
    color = "black"# color of bin outline 
    ) 
```

## Changing the color of the bars

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             fill = gender) # note: using "fill" not "color"
       ) + 
  geom_histogram(# adds the histograms to the graph
    bins = 10, # number of bins 
    color = "black"# color of bin outline 
    ) +
  scale_fill_manual(values = c("female" = 'lightgrey', # sets colors for each level
                                "male" = 'darkgrey'))
```

## Modifying plot titles, labels, and theme

```{r}
ggplot(data = covid_testing, # dataset to use for plot
                   mapping = aes( # list of aesthetic mappings to use for plot
                             x = pan_day, # x-axis variable
                             fill = gender) # note: using "fill" not "color"
       ) + 
  geom_histogram(# adds the histograms to the graph
    bins = 10, # number of bins 
    color = "black"# color of bin outline 
    ) +
  scale_fill_manual(values = c("female" = 'lightgrey', # sets colors for each level
                                "male" = 'darkgrey')) + 
  ggtitle("Number of Patients by Pandemic Day") + # adds title to plot
  xlab("Pandemic Day") + # changes x-axis label
  ylab("Number of Patients") + # changes y-axis label 
  labs(fill = "Patient Gender") + # changes legend label (note this says "fill")
  theme_bw() + # changes plot theme to black and white
  theme(plot.title = element_text(hjust = 0.5), # centers plot title
        legend.position="bottom") # moves legend to bottom 
```

## CODING CHALLENGE 3

**Your turn!**

Using the same covid_testing dataset, use ggplot to recreate the "Testing Results by Patient Age" figure in the presentation. 

You can also try saving the figure to a directory of your choice.

Reminder: Codebook for the covid_testing dataset is available [here](https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/codebooks/covid_testing_codebook.html)

```{r}




```

There are lots of examples of things to try here 
https://r-graph-gallery.com/ggplot2-package.html

** ---------------------------------------------- **
** THE END!  Thank you for participating!         **
** ---------------------------------------------- **
