## Load packages

If you're running this on your own computer rather than in the cloud instance we set up, you may need to run `install.packages("tidyverse")` first if you haven't ever used the tidyverse packages before.

```{r}
library(tidyverse)
```


## The data

```{r}
head(msleep)
```

To learn more about this dataset:

```{r}
?msleep
```

## How to check for missing values

```{r}
summary(msleep)
```

## Filtering with `NA`

Let's use `filter()` to take a look at just the rows that have missing values for the `brainwt` variable.

```{r}
# This doesn't work!
filter(msleep, brainwt == NA)
```


```{r}
filter(msleep, is.na(brainwt))
```




## CODING CHALLENGE 1!

Now fill in the partial code below to filter the data to just look at rows that are **not** missing for `brainwt`. 

```{r}
filter(msleep, !is.na(brainwt))
```

When you're done, put a "thumbs up" on the note in the chat to let us know you're ready to move forward. :)

----------------------------------------------------------







## CODING CHALLENGE 2!

Let's pretend you know that these data were only recorded over an 18 hour period, so it would have been impossible to have a total sleep time of greater than 18. And yet, when we look at the data, we see there are a few animals with `sleep_total` over 18:

```{r}
# summary of the sleep_total variable
summary(msleep$sleep_total)
```

```{r}
# show just the rows of msleep where sleep_total is greater than 18
filter(msleep, sleep_total > 18)
```

These must be typos or some other kind of error, so you want to exclude them from your analysis.

Fill in this code to mark values for `sleep_total` that are over 18 as missing (remember you can look at the solutions file if you get stuck!):

```{r}
msleep <- mutate(msleep, 
                 sleep_total = ifelse(sleep_total > 18, NA, sleep_total))
```

To check that it worked, take a look at the updated summary stats for `sleep_total`:

```{r}
summary(msleep$sleep_total)
```

You should see that the Max value is now no greater than 18, and you should see some NAs.


When you're done, put a "thumbs up" on the note in the chat to let us know you're ready to move forward. :)

----------------------------------------------------------




Use this space to run tests and try things out!

```{r}

```


