---
title: "Summary Statistics"
output: github_document
---

## Load packages

```{r}
library(tidyverse)
library(medicaldata) 
library(gtsummary)
```

## The data

```{r}
head(cytomegalovirus)
```


```{r}
# To learn more about this dataset
?cytomegalovirus
```

## Using `summary()`

```{r}
# convert diagnosis column from character to factor
cytomegalovirus$diagnosis <- as.factor(cytomegalovirus$diagnosis)

# re-run the summary for that column
summary(cytomegalovirus$diagnosis)
```

```{r}
# clean up variables that should be factors
cytomegalovirus <- cytomegalovirus |> 
  mutate(sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
         race = factor(race, levels = c(0, 1), labels = c("African-American", "White")),
         diagnosis.type = factor(diagnosis.type, levels = c(0, 1), labels = c("Lymphoid", "Myeloid")),
         prior.radiation = factor(prior.radiation, levels = c(0, 1), labels = c("No", "Yes")),
         prior.transplant = factor(prior.transplant, levels = c(0, 1), labels = c("No", "Yes")),
         donor.cmv = factor(donor.cmv, levels = c(0, 1), labels = c("Negative", "Positive")),
         recipient.cmv = factor(recipient.cmv, levels = c(0, 1), labels = c("Negative", "Positive")),
         donor.sex = factor(donor.sex, levels = c(0, 1), labels = c("Male", "Female")),
         `C1/C2` = factor(`C1/C2`, levels = c(0, 1), labels = c("Heterozygous", "Homozygous")),
         cmv = factor(cmv, levels = c(0, 1), labels = c("No", "Yes")),
         agvhd = factor(agvhd, levels = c(0, 1), labels = c("No", "Yes")),
         cgvhd = factor(cgvhd, levels = c(0, 1), labels = c("No", "Yes"))
         )

summary(cytomegalovirus)
```


## Individual summary statistics

```{r}
min(cytomegalovirus$time.to.transplant)

min(cytomegalovirus$time.to.transplant, na.rm = TRUE)

max(cytomegalovirus$time.to.transplant, na.rm = TRUE)

mean(cytomegalovirus$time.to.transplant, na.rm = TRUE)

median(cytomegalovirus$time.to.transplant, na.rm = TRUE) # median
var(cytomegalovirus$time.to.transplant, na.rm = TRUE) # variance
sd(cytomegalovirus$time.to.transplant, na.rm = TRUE) # standard deviation

# the default quantiles
quantile(cytomegalovirus$time.to.transplant, na.rm = TRUE) 

# custom quantiles
quantile(cytomegalovirus$time.to.transplant, probs = c(.1, .2, .3, .4), na.rm = TRUE)
```

## CODING CHALLENGE 1

**Your turn!**

What is the average (mean) dose for the following cells: TNC, CD34, CD3, and CD8? (If you're not sure what the column names are for those variables, check the [code book for this dataset](https://www.causeweb.org/tshs/datasets/Cytomegalovirus%20Data%20Dictionary.pdf))

```{r}
# mean dose for TNC
mean(cytomegalovirus$TNC.dose, na.rm = TRUE)
# mean dose for CD34
mean(cytomegalovirus$CD34.dose, na.rm = TRUE)
# mean dose for CD3
mean(cytomegalovirus$CD3.dose, na.rm = TRUE)
# mean dose for CD8
mean(cytomegalovirus$CD8.dose, na.rm = TRUE)
```

Now use `summary()` to get the min, max, median, mean, 25th percentile, 75th percentile, and missingness count for all of the cell dose variables (TNC, CD34, CD3, and CD8). 
Hint: try using `select()` to get just the variables you want to work with. 

```{r}
# use summary() to get the min, max, median, mean, 25th percentile, 75th percentile, and missingness count for all of the cell dose variables (TNC, CD34, CD3, and CD8)
cytomegalovirus |> 
  select(TNC.dose, CD34.dose, CD3.dose, CD8.dose) |> 
  summary()

```

** ---------------------------------------------- **
** STOP HERE!  Go back to the module to continue. **
** ---------------------------------------------- **

## Counts

```{r}
# two ways to get the counts for a single categorical variable
summary(cytomegalovirus$diagnosis.type)
table(cytomegalovirus$diagnosis.type)

# how to see missing values with table
table(cytomegalovirus$diagnosis.type, useNA = "always")
```

```{r}
# counts of two variables at once (cross-tablulation)
xtabs(~cmv + diagnosis.type, data = cytomegalovirus)
```

```{r}
# two ways to get the results of a chi-squared test on the crosstab:

# you can nest the summary and xtabs functions
summary(xtabs(~cmv + diagnosis.type, data = cytomegalovirus))

# or you can save the xtabs object and then run summary() on it
cmv_and_diagnosistype <- xtabs(~cmv + diagnosis.type, data = cytomegalovirus)
summary(cmv_and_diagnosistype)
```

## CODING CHALLENGE 2

**Your turn!** 

Get the counts for both recipient cmv status and donor cmv status, separately. Then generate a crosstab for recipient cmv status and donor cmv status, and get the corresponding chi-squared test of independence. (Hint: If you're not sure what the column names are for those two variables, check the [code book for this dataset](https://www.causeweb.org/tshs/datasets/Cytomegalovirus%20Data%20Dictionary.pdf)) or use `?cytomegalovirus`.

```{r}
# get the counts for recipient cmv status
summary(cytomegalovirus$cmv)

# get the counts for donor cmv status
summary(cytomegalovirus$donor.cmv)
```

```{r}
# make a crosstab for recipient cmv status and donor cmv status
cmv_xtab <- xtabs(~cmv + donor.cmv, data = cytomegalovirus)

# get the chi-squared test of independence for the relationship between recipient cmv status and donor cmv status
summary(cmv_xtab)

cmv_xtab # print out the xtab table itself
```

** ---------------------------------------------- **
** STOP HERE!  Go back to the module to continue. **
** ---------------------------------------------- **

## Creating Table 1 with `gtsummary`

```{r}
# first we need to create a factor version of aKIRs, breaking it down as aKIRs 1-4 or aKIRs 5-6
# we also need to break prior.chemo into a factor with levels 0, 1, 2, 3, and ≥ 4
cytomegalovirus <- cytomegalovirus |> 
  mutate(aKIRs_groups = if_else(aKIRs > 4, "5 to 6 aKIRs", "1 to 4 aKIRs"),
         aKIRs_groups = factor(aKIRs_groups, levels = c("1 to 4 aKIRs", "5 to 6 aKIRs")),
         prior.chemo_groups = if_else(prior.chemo >= 4, "≥ 4", as.character(prior.chemo)),
         prior.chemo_groups = factor(prior.chemo_groups, levels = c("0", "1", "2", "3", "≥ 4")))
```

  
```{r}  
cytomegalovirus |>  
  select(aKIRs_groups, age, sex, race, diagnosis.type, time.to.transplant, prior.radiation, prior.chemo_groups, prior.transplant, recipient.cmv, donor.cmv) |> 
  tbl_summary(by = aKIRs_groups)

# to save re-typing code and highlight just what's new, save the first part
table_data <- cytomegalovirus |>  
  select(aKIRs_groups, age, sex, race, diagnosis.type, time.to.transplant, prior.radiation, prior.chemo_groups, prior.transplant, recipient.cmv, donor.cmv)
```


```{r}
# add a column showing p values
table_data |> 
  tbl_summary(by = aKIRs_groups) |> 
  add_p() 
```

```{r}
# make the p values round to 2 digits instead of 1, and use t.test instead of wilcoxon rank sum test
table_data |> 
  tbl_summary(by = aKIRs_groups) |> 
  add_p(test = list(all_continuous() ~ "t.test",
                    all_categorical() ~ "chisq.test.no.correct"),
        test.args = all_continuous() ~ list(var.equal = TRUE),
        pvalue_fun = function(x) style_pvalue(x, digits = 2)) 
```

```{r}
# customize the row labels
table_data |> 
  tbl_summary(by = aKIRs_groups, 
              label = list(age ~ "Age at transplant (yrs)",
                           sex ~ "Sex",
                           race ~ "Race",
                           diagnosis.type ~ "Diagnostic category",
                           time.to.transplant ~ "Months from diagnosis to transplant",
                           prior.radiation ~ "Prior radiation therapy",
                           prior.chemo_groups ~ "Number of prior chemotherapy regimens",
                           prior.transplant ~ "Prior transplants",
                           recipient.cmv ~ "Recipient CMV seropositive",
                           donor.cmv ~ "Donor CMV seropositive")) |> 
  add_p(test = list(all_continuous() ~ "t.test",
                    all_categorical() ~ "chisq.test.no.correct"),
        test.args = all_continuous() ~ list(var.equal = TRUE),
        pvalue_fun = function(x) style_pvalue(x, digits = 2)) 

```

## CODING CHALLENGE 3

**Your turn!**

Let's try creating Table 1 for a new dataset. We'll use another dataset that's available through the `medicaldata` package, so there's no need to install or import anything else. The data is called `laryngoscope` and it's from a randomized control trial comparing two different laryngoscopes.

Check out the [published article on the laryngoscope data](https://pubmed.ncbi.nlm.nih.gov/21918156/) for an overview. Here's Table 1:

![Table 1 from the laryngoscope paper.](https://journals.lww.com/anesthesia-analgesia/_layouts/15/oaks.journals/ImageView.aspx?k=anesthesia-analgesia:2011:11000:00024&i=T1-24&year=2011&issue=11000&article=00024)

Before you start writing code, you'll need to look over the description of the variables in the data, so you know what's what. That's all available in the [codebook](https://www.causeweb.org/tshs/datasets/Laryngoscope%20Data%20Dictionary.pdf), or in the help documentation for the data (run `?laryngoscope` in the console).


```{r}
# Here's a first attempt
laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization)
```

That's a perfectly respectable table! There are lots of options for customization with tbl_summary, though, so we'll show a few tweaks that bring this table closer to the publication version.

```{r}
# Let's change the Randomization, gender, and asa variables to have descriptive labels
# and make Mallampati a factor since it's categorical, but no need to assign labels (it should just print with numbers)
laryngoscope <- laryngoscope |> 
  mutate(Randomization = factor(Randomization, levels = c(1, 0), labels = c("Pentax AWS", "Macintosh")),
         gender = factor(gender, levels = c(0, 1), labels = c("female", "male")),
         asa = factor(asa, levels = c(2, 3, 4), labels = c("II", "III", "IV")),
         Mallampati = as.factor(Mallampati))

laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization)
```

```{r}
# clean up the labels
laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization,
              label = list(age ~ "Age, y",
                           gender ~ "Gender",
                           BMI ~ "Body Mass Index, kg/m2",
                           asa ~ "ASA physical status",
                           Mallampati ~ "Mallampati score"))
```


```{r}
# change the stats displayed to be mean +- SD for continuous variables and number (%) for categorical
# remove the rows showing missing observations
laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization,
              label = list(age ~ "Age, y",
                           gender ~ "Gender",
                           BMI ~ "Body Mass Index, kg/m2",
                           asa ~ "ASA physical status",
                           Mallampati ~ "Mallampati score"),
              statistic = list(all_continuous() ~ "{mean} ± {sd}",
                               all_categorical() ~ "{n} ({p})"),
              missing = "no") 
```

```{r}
# add a column for the standardized difference
laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization,
              label = list(age ~ "Age, y",
                           gender ~ "Gender",
                           BMI ~ "Body Mass Index, kg/m2",
                           asa ~ "ASA physical status",
                           Mallampati ~ "Mallampati score"),
              statistic = list(all_continuous() ~ "{mean} ± {sd}",
                               all_categorical() ~ "{n} ({p})"),
              missing = "no") |> 
  add_difference(test = all_continuous() ~ "cohens_d")
```


```{r}
# hide the CI column
laryngoscope |> 
  select(Randomization, age, gender, BMI, asa, Mallampati) |> 
  tbl_summary(by = Randomization,
              label = list(age ~ "Age, y",
                           gender ~ "Gender",
                           BMI ~ "Body Mass Index, kg/m2",
                           asa ~ "ASA physical status",
                           Mallampati ~ "Mallampati score"),
              statistic = list(all_continuous() ~ "{mean} ± {sd}",
                               all_categorical() ~ "{n} ({p})"),
              missing = "no") |> 
  add_difference(test = all_continuous() ~ "cohens_d") |> 
  modify_column_hide(ci)
```


Keep playing around with the formatting for the table! There are lots of examples of things to try here https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html 

** ---------------------------------------------- **
** STOP HERE!  Go back to the module to continue. **
** ---------------------------------------------- **
